Maximum Entropy Spectral Analysis: a case study

### 2024-02-12 paper/mesa.tex
final paper draft

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TITLE
	\title{Maximum Entropy Spectral Analysis: a case study}
	\author{Alessandro \surname{Martini}\inst{\ref{inst1}}\thanks{martini.alessandr@gmail.com} \and 
	Stefano \surname{Schmidt}\inst{\ref{inst2}}\thanks{s.schmidt@uu.nl} \and Walter \surname{Del Pozzo}\inst{\ref{inst1}}\thanks{walter.delpozzo@unipi.it}}

@@

  \abstract{
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on a Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum}. We characterize our implementation by performing a power spectral density analysis on synthetic data (with known power spectral density) and we compare different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using synthetic data generated from the released spectrum by the LIGO-Virgo-Kagra collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance and bias. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime. 
We conclude our investigation by utilising MESA, and its particularly easy parametrisation where the only free parameter is the order $p$ of the AR process, to marginalise over the interferometers noise PSD in conjunction with inferring the parameters of GW150914.}

@@

\section{Introduction}
\todo{Check again intro \& abstract}
The study of the properties of stochastic processes is a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among those, a special place is occupied by the so-called \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to time translation. If $x(t)$ is a wide-sense stationary process, it is completely determined by the knowledge of the autocorrelation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, by the knowledge of their \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem the two are related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In some literature, especially in the context of gravitational waves physics, e.g. ~\citep{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

An ubiquitous method for such computation is due to \citet{Welch1967} and it is based on Eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
The PSD is obtained by slicing the observed realization $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many window-corrected batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent~\citep{Lomb,Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different ways, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

A alternative approach providing a smooth PSD estimation, is to adopt  a parametric model for the PSD and to fit its parameters to the data with a Markov Chain Monte Carlo \citep[e.g.]{Cornish_2015, Littenberg_2015}. Despite being effective, this method is problem dependent, since it needs to make definite assumptions on the shape of the PSD. Moreover, it can be computationally expensive and it does not come with an handy implementation available to the public. For all the above reasons, we did not consider such method in our work.

An appealing alternative, based on the Maximum Entropy principle \citep{JaynesArticle,jaynes2003ptl, Jaynes_MAXENT}, has been put forward by~\citet{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the algorithm parameters, since it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provides a natural and simple machinery to forecast a time series, thus predicting future observations based on previous ones.

In this paper, we discuss the details of the Maximum entropy principle, its application to the problem of PSD estimation with Burg's algorithm and the link between Burg's algorithm and autoregressive process.
Our goal is to bring (again) to public attention Maximum Entropy Spectral analysis, in the hope that it will be widely employed as a way out to the many undesired aspects of the Welch's algorithm (or other similar methods).
To facilitate this goal, we present and describe a new code, \texttt{memspectrum}, that provides a robust and easy-to-use python implementation of the algorithm\footnote{
It is available at link: \url{https://pypi.org/project/memspectrum/}.
}.
We provide a thorough assessment of the performance of our code and we validate our results performing a number of tests on simulated and real data.
We also compare our results with those of spectral analysis carried out with the standard Welch's method.
In order to apply our model on a realistic setting, we analyse some time series of broad interest in the scientific community.

@@

\section{Theoretical foundations}\label{sec:foundations}
The Maximum Entropy principle (MAXENT) is among the most important results in probability theory. It provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while being non committal with unavailable information. Its domain of application turned out to be wider than expected. In fact, thanks to~\citet{burg1975maximum}, this method has also been applied to perform high quality computation of power spectral densities of time series.

After a short introduction to Jaynes' MAXENT (Sec.~\ref{sec:MAXENT}), we will develop in detail Burg's technique of Maximum Entropy Spectral Analysis (MESA) and show that the estimate can be expressed in an analytical closed form (Sec.~\ref{sec:MESA}).
Next, we will discuss an interesting link between Burg's method and autoregressive processes (Sec.~\ref{sec:autoregr}) and in Sec.~\ref{sec:forecasting} we will use such link for straightforwardly forecasting a time series.

@@

Before introducing MAXENT principle, we will define through some simple examples the two core concepts of the problem and the role they play: the `evidence' and the `information'.
Let us start with the `information' (or entropy): it is a measure of the degree of uncertainty on the outcomes of some experiment and specifies the length of the message necessary to provide a full description of the system under study. For instance, consider a perfectly sinusoidal signal: knowledge of amplitude, frequency and phase are sufficient to fully reproduce it: it has a low information content. Even less information is required if we are studying a system whose outcome is certain (has probability $p = 1$), as in this case, a communication is not even needed.  
\citet{Shannon} proposed the quantity
\begin{equation}\label{eq:information}
    I = \log_2 \frac{1}{p(x)}
\end{equation}
to measure the quantity of information brought by an outcome $x$ with probability $p(x)$. It is additive quantity as well as monotonically decreasing as a function of $p \in [0, 1]$: the more uncertain the outcome, the higher the information it brings.

We can generalize the definition of information in the case where two different outcomes $E_1, E_2$, with given probabilities $P_1$ and $P_2$, are possible.
To gain some intuition on the problem, we ask ourselves which are the probability assignments that make the outcome more uncertain (i.e. maximize the information).
If $P_1$ and $P_2$ are largely different, for instance $P_1 = 0.999$ and $P_2 = 0.001$, we are allowed to believe that event $E_1$ will occur almost certainly, considering $E_2$ to be a very implausible outcome. The information content will be very low.
On the other hand, most unpredictable experiment happens when 
\begin{equation}\nonumber
    P_1 = P_2 = \frac{1}{2}:
\end{equation}
this describes a situation of `maximum ignorance' and the information content of such system must be high.
Any generalization of eq.~\eqref{eq:information}, must then have its maximum when $ P_1 = P_2$.
For $N$ events, the system with the highest possible information content is when:
\begin{equation}\nonumber
    P_1 = \hdots = P_N = \frac{1}{N}:
\end{equation}

\citet{Shannon} showed that the only functional form satisfying continuity with respect to its parameters, additivity and that has a maximum for equal probability events is:
\begin{equation}\label{eq:entropy}
    H[p_1, \dots, p_N] = - \sum_{i = 1}^N p_i\log{p_i},
\end{equation}
The equation can be interpreted as the `expected information' brought by an experiment with $N$ possible outcomes each with its own probability $p_i$.
In the continuous case:
\begin{equation} \label{eq:entropy_continuos}
    H[p(x)] = - \int p(x)\ln p(x) dx,
\end{equation}
We call the functional $H$ (information) entropy\footnote{In defining the information entropy as in Eq.~(\ref{eq:entropy_continuos}) we are implicitly assuming a uniform measure over the parameter space}.

We now turn to the core of our problem: how do we make a probability assignment for a set of events, that keeps into account our knowledge of the system and, at the same time, it is non committal towards unavailable knowledge?
The knowledge at our disposal about the system under study is what we call `evidence' and any probability assignment must agree with it.
In the above cases, our knowledge on the system is only the total number $N$ of different outcomes -- this is a minimal requirement. Of course, more complex evidence constraints can be applied.

It is very common that the constraints provided by the evidence are not enough for setting the probabilities for each event: in this case, it is reasonable to assume that the probability assignment should make the experiment as unpredictable as possible\footnote{
In~\citet{Jaynes_MAXENT} this statement is made more precise and justified more thoroughly, with arguments based on combinatorial analysis.
}.
In other words, the amount of `information' introduced by the probability assignment should be as high as possible.

MAXENT puts the aforementioned principle into a more precise mathematical context by stating that probabilities should be assigned by maximizing uncertainty (information entropy) using evidence as constraint. 
This defines a variational problem, where the information entropy functional $H\left[p_1, \dots, p_N\right]$, defined in eq.~\eqref{eq:entropy}, has to be maximized.

The maximisation of entropy, supplemented by evidence in the form of constraints to which the sought-for probability distribution must obey,
gives rise to several of the most common probability distributions commonly employed in statistics. For instance, whenever the only constraint available is the normalization of the probability distribution (i.e. no evidence is available), the entropy is maximised by the uniform distribution. 
If we have evidence to constraint the expected value, the information entropy is maximised by the exponential distribution

Of particular importance is the case in which, in addition to the mean, also the variance is known: MAXENT leads to the Gaussian distribution. 
This derivation is particularly interesting from the foundational point of view, since it provides a deeper insight into the ubiquitous Gaussian distribution.
Indeed, it is not only the limit distribution provided by the central limit theorem for finite variance processes but it is also the distribution that maximizes the entropy. For this reason, appealing to MAXENT principle, it is the correct assignment if mean and covariance are the only quantities that fully define our process. In some sense, we can interpret the central limit theorem as the natural `statistical' evolution toward a configuration that maximizes entropy.

For this work, we are particularly interested in the multi-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realization of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of autocovariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution~\citep{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

For a wide-sense stationary process the mean function is independent of time, hence it can be redefined to be equal to zero without loss of generality, and the auto-covariance function is dependent only on the time lag $\tau \equiv t_i - t_j$. One can thus choose a sampling rate $\Delta t$ so that $C_{ij} = C((i-j)\Delta t)$. The autocovariance matrix thus becomes a Toeplitz matrix\footnote{
We remind the reader that a Toeplitz matrix is a matrix in the form:
\begin{pmatrix}
	a_0 & a_1 & a_2 & \ldots & \ldots& \ldots &a_n\\
	a_{-1} & a_0 & a_1 & \ldots & \ldots& \ldots &a_{n-1}\\
	a_{-2} & a_{-1} & a_0 & \ldots & \ldots & \ldots  &a_{n-2}\\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots &\vdots\\
	a_{-n +1} & \ldots & \ldots & \ldots& a_{-1} & a_0    &a_{1}\\
	a_{-n} & \ldots & \ldots & \ldots& a_{-2} & a_{-1} & a_0
\end{pmatrix}
}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalized by the discrete Fourier transform base~\citep{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_{n/2})\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{n / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} \tilde{x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognize the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\citep{prob_information_theory}
and of gravitational waves data analysis, \citep[e.g.]{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the autocovariance function.

@@

In principle, if the autocorrelation was known exactly (i.e. at every time $\tau \in (-\infty,+\infty)$), the computation of the PSD 
would reduce to a single Fourier transform.
However, in any realistic setting, we are dealing with a finite number of samples $N$ from the process, hence such computation
is impossible.
Moreover, the error $\sigma_k$ in the estimate of the autocorrelation after $k$ steps increases as $\sigma \sim 1/\sqrt{N - k}$\footnote{
This is clearly understood: when computing the autocorrelation at order $k$, only $N-k$ examples of the product $x_t x_{t+k}$ are available and the variance of the average value goes as the inverse of the square root of the points considered.
}, so that only few values for the autocorrelation function can actually be computed reliably.
This bring us the core of the problem: how to give an estimate from partial (and noisy) knowledge of the autocorrelation function? MAXENT can guide us in this task without any a priori assumptions on the unavailable data\footnote{Indeed this is the largest difference with the most common Welch method. The latter assumes that the unknown values of the autocorrelation are $0$. Clearly, this assumption is unjustified and MAXENT is a good way to drop this assumption.}.

@@

Before doing so, there is a technicality to solve: the definition of entropy depends on a probability distribution, not on the PSD.
It can be shown, ~\citep[e.g.]{AblesMESA, Bartlett}, that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of the filtering a white noise process using a filter with transfer function $T(f)$ equal to $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising Eq.~\eqref{eq:EntropyGain} is equivalent to maximizing eq.~\eqref{eq:entropy_continuos}.

>>>>

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% TITLE
	\title{Maximum Entropy Spectral Analysis: a case study}
	\author{Alessandro \surname{Martini}\inst{\ref{inst1}}\thanks{martini.alessandr@gmail.com} \and 
	Stefano \surname{Schmidt}\inst{\ref{inst2}}\thanks{s.schmidt@uu.nl} \and Gregory \surname{Ashton}\inst{\ref{inst1}}\thanks{walter.delpozzo@unipi.it} \and Walter \surname{Del Pozzo}\inst{\ref{inst1}}\thanks{walter.delpozzo@unipi.it}}

@@

  \abstract{
The Maximum Entropy Spectral Analysis (MESA) method, developed by Burg, provides a powerful tool to perform spectral estimation of a time-series. The method relies on Jaynes' maximum entropy principle and provides the means of inferring the spectrum of a stochastic process in terms of the coefficients of some autoregressive process AR($p$) of order $p$.
A closed form recursive solution provides an estimate of the autoregressive coefficients as well as of the order $p$ of the process.
We provide a ready-to-use implementation of the algorithm in the form of a python package \texttt{memspectrum} and we characterize our implementation by performing a power spectral density analysis on synthetic data (with known power spectral density) and by comparing different criteria for stopping the recursion. Furthermore, we compare the performance of our code with the ubiquitous Welch algorithm, using synthetic data generated from the released spectrum by the LIGO-Virgo-Kagra collaboration.
We find that, when compared to Welch's method, Burg's method provides a power spectral density (PSD) estimation with a 
systematically lower variance and bias. This is particularly manifest in the case of a little number of data points, making Burg's method most suitable to work in this regime. 
We conclude our investigation by utilising MESA, and its particularly easy parametrisation where the only free parameter is the order $p$ of the AR process, to marginalise over the interferometers noise PSD in conjunction with inferring the parameters of GW150914.}

@@

\section{Introduction}

The problem of inferring the morphology and the defining parameters of deterministic signals superimposed to stochastic processes is one of the most wide spread and interesting problems in several areas of human activities. Whenever some form of model for the signal we are looking for is available, the problem is typically solved via the Wiener filter, defined as the whitening filter that maximises the signal-to-noise ratio, i.e. the relative power of the (known) signal over the power of the (known) underlying stochastic process. Hence, signal detection and characterisation requires accurate knowledge of i) the shape of the signal we are looking for and ii) the statistical properties of the stochastic process. The construction of signal models is typically driven either by physical or by mathematical arguments hence, although extremely difficult in general, it is doable. On the other hand, stochastic process models can be extremely difficult to construct, both for practical and theoretical reasons. A stochastic process is fully described by the knowledge of the probability distribution governing its realisations -- the ``paths'' of the random variable under scrutiny -- over the entire time axis, from $t=-\infty$ to $t=\infty$. Clearly this is not possible in practice. Therefore modelling a stochastic process either relies on modelling of the underlying physical processes, thus falling back onto the deterministic case, or on modelling the mathematical and statistical properties of the process, and potentially infer them from the process realisations. The study of the properties of stochastic processes is thus a crucial task in many fields of physics, astronomy, quantitative biology, as well as engineering and finance. Among the classes of stochastic processes, a key role is played by \textit{wide-sense} stationary processes. These are stochastic processes that display an invariance of their statistical properties, such as their two-point autocovariance function, with respect to translation of the independent variable, usually the time $t$. If $x(t)$ is a wide-sense stationary process, it is in fact completely determined by the knowledge of the (many-points) autocorrelation functions. In practice, one often has easy access to the two-point correlation function 
\begin{equation}
	C(\tau) = \mathbf{E}[x_t \cdot x_{t+\tau}]
\end{equation}
or, equivalently, to the process \emph{power spectral density} (PSD) $S(f)$. Thanks to the Wiener-Khinchin theorem, in wide-sense stationary processes, the two are in fact related by a Fourier transform: 
\begin{equation}\label{eq:psd-autocorrelation}
	S(f) = \int_{-\infty}^{\infty} \textrm{d}\tau C(\tau) e^{-i 2 \pi f \tau}\,.
\end{equation}
In the context of gravitational waves physics, e.g. ~\citep{Finn_1992}, the PSD is introduced as  
\begin{equation}\label{eq:psd-f-definition}
	\mathbf{E}[\tilde{x}(f) \cdot \tilde{x}(f^\prime)] = S(f) \delta(f-f^\prime)
\end{equation}
without highlighting its connection with the time structure of the process itself, thus masking some important properties that will be explored further in what follows. The latter definition in~\eqref{eq:psd-f-definition} gives, however, i) a straightforward interpretation of the PSD: it measures how much signal ``power" is located in each frequency; ii) an operative way of estimating it for an unknown process.

An ubiquitous method for such a computation is due to \citet{Welch1967} and it is based on Eqs.(\ref{eq:psd-autocorrelation}-\ref{eq:psd-f-definition}).
The PSD is obtained by slicing the observed realisation $x(t_1),\ldots,x(t_n)$ of the process $x(t)$ into many window-corrected batches and averaging the squared moduli of their Fourier transforms.
This approach is equivalent~\citep{Lomb,Scargle} to taking the Fourier Transform of the windowed sample autocorrelation $\rho_W$, written as
\begin{equation}
    \rho_{W} = \left\{W_0\rho_0,W_{\pm 1}\rho_{\pm 1}, \dots, W_{\pm M}\rho_{\pm M}, 0, 0, \dots \right\},
\end{equation}
where $\rho$ is the empirical autocorrelation and $M$ is the maximum time lag at which the autocorrelation is computed.
The sequence $W$ is a window function that can be chosen in several different ways, each choice presenting advantages and disadvantages for the final estimate of the PSD.

@@

A alternative approach providing a smooth PSD estimation, is to adopt a parametric model for the PSD and to fit its parameters to the data with a Markov Chain Monte Carlo \citep[e.g.]{Cornish_2015, Littenberg_2015}. Despite being effective, this method is problem dependent, since it needs to make definite assumptions on the shape of the PSD. Moreover, it can be computationally expensive and it does not come with a handy implementation available to the public. For all the above reasons, we did not consider such methods in our work.

An appealing alternative, based on the Maximum Entropy principle \citep{JaynesArticle,jaynes2003ptl, Jaynes_MAXENT}, has been derived by~\citet{burg1975maximum}. Being rooted on solid theoretical foundations, we will see that Burg's method, unlike Welch's, does not require any preprocessing of the data and requires very little tuning of the algorithm parameters, since it provides an iterative closed form expression for the spectrum of a stochastic stationary time series. Furthermore, it embeds the PSD estimation problem into an elegant theoretical framework and makes minimal assumptions on the nature of the data.
Lastly and most importantly, it provides a robust link between spectral density estimation and the field of autoregressive processes. This provides a natural and simple machinery to forecast a time series, thus predicting future observations based on previous ones.

In this paper, we discuss the details of the Maximum entropy principle, its application to the problem of PSD estimation with Burg's algorithm and the link between Burg's algorithm and autoregressive process.
Our goal is to bring (again) to public attention Maximum Entropy Spectral analysis, in the hope that it will be widely employed as a way out to the many undesired aspects of the Welch's algorithm (or other similar methods).
To facilitate this goal, we based this study on \texttt{memspectrum}, a freely available, robust and easy-to-use python implementation of the algorithm described below\footnote{
It is available at link: \url{https://pypi.org/project/memspectrum/}.}.
We provide a thorough assessment of the performance of our code and we validate our results performing a number of tests on simulated and real data.
We also compare our results with those of spectral analysis carried out with the standard Welch's method.
In order to apply our model on a realistic setting, we analyse some time series of broad interest in the scientific community.

@@

\section{Theoretical foundations}\label{sec:foundations}
The Maximum Entropy principle (MAXENT) is among the most important results in probability theory. It provides a way to uniquely assign probabilities to a phenomenon in a way that best represent our state of knowledge, while being non-committal to unavailable information. Its domain of application turned out to be wider than expected. In fact, thanks to~\citet{burg1975maximum}, this method has also been applied to perform high quality computation of power spectral densities of time series.

After a short introduction to Jaynes' MAXENT (Sec.~\ref{sec:MAXENT}), we will review in detail Burg's technique of Maximum Entropy Spectral Analysis (MESA) and show that the estimate can always be expressed in an analytical closed form (Sec.~\ref{sec:MESA}).
Next, we will discuss the interesting link between Burg's method and autoregressive processes (Sec.~\ref{sec:autoregr}) and in Sec.~\ref{sec:forecasting} we will use such link to forecast a time series.

@@

Before introducing MAXENT principle, we will define, via some simple examples, the two core concepts of the problem and the roles they play in deductive inference: the `evidence' and the `information'.
Let us start with the `information' (or information entropy): it is a measure of the degree of uncertainty on the outcomes of some experiment and specifies the length of the message necessary to provide a full description of the system under study. For instance, consider a perfectly sinusoidal signal: knowledge of amplitude, frequency and phase are sufficient to fully reproduce it: it has a low information content. Even less information is required if we are studying a system whose outcome is certain (the outcome is known with probability $p = 1$), as in this case, a communication is not even needed.  
\citet{Shannon} proposed the quantity
\begin{equation}\label{eq:information}
    I = \log_2 \frac{1}{p(x)}
\end{equation}
to measure the quantity of information brought by an outcome $x$ with probability $p(x)$. It is additive quantity as well as a monotonically decreasing function of $p \in [0, 1]$: the more uncertain the outcome, the higher the information it brings.

We can generalize the definition of information in the case where two different outcomes $E_1, E_2$, with given probabilities $P_1$ and $P_2$, are possible.
To gain some intuition on the problem, we ask ourselves which are the probability assignments that make the outcome more uncertain (i.e. maximize the information).
If $P_1$ and $P_2$ are largely different, for instance $P_1 = 0.999$ and $P_2 = 0.001$, we are allowed to believe that event $E_1$ will occur almost certainly, considering $E_2$ to be a very implausible outcome. The information content will be very low.
On the other hand, most unpredictable situation happens when 
\begin{equation}\nonumber
    P_1 = P_2 = \frac{1}{2}:
\end{equation}
this describes a situation of `maximum ignorance' and the information content of such system must be high.
Any generalization of Eq.~\eqref{eq:information}, must then have its maximum when $ P_1 = P_2$.
For $N$ events, the system with the highest possible information content is when:
\begin{equation}\nonumber
    P_1 = \hdots = P_N = \frac{1}{N}:
\end{equation}

\citet{Shannon} showed that the only functional form satisfying continuity with respect to its parameters, additivity and that has a maximum for equal probability events is:
\begin{equation}\label{eq:entropy}
    H[p_1, \dots, p_N] = - \sum_{i = 1}^N p_i\log{p_i}\,,
\end{equation}
which can be interpreted as the `expected information' brought by an experiment with $N$ possible outcomes each with its own probability $p_i$.
In the continuous case:
\begin{equation} \label{eq:entropy_continuos}
    H[p(x)] = - \int p(x)\ln p(x) dx,
\end{equation}
We call the functional $H$ (information) entropy\footnote{In defining the information entropy as in Eq.~\eqref{eq:entropy_continuos}, we are implicitly assuming a uniform measure over the parameter space. In case of a non-uniform measure $m(x)$, the definition generalises to $ H[p(x)] = - \int p(x)\ln \frac{p(x)}{m(x)} dx$.}.

We now turn to the core of our problem: how can we assign probabilities to a set of events keeping into account our knowledge of the system and, at the same time, it is non-committal towards unavailable knowledge?
The ``knowledge'' at our disposal about the system under investigation is what we define `evidence' and any probability assignment is given such evidence, in agreement with \citet{Cox} construction of probability. In the case above, our knowledge on the system is only the total number $N$ of different outcomes -- this is a minimal requirement. Of course, more complex evidence constraints can be applied.

It is very common that the constraints provided by the evidence are not enough for setting the probabilities for each event: in this case, it is reasonable to assume that the probability assignment should make the experiment as unpredictable as possible\footnote{
In~\citet{Jaynes_MAXENT} this statement is made more precise and justified more thoroughly, with arguments based on combinatorial analysis.
}.
In other words, the information entropy content introduced by the probability assignment should be as large as possible, in accordance with the available evidence. MAXENT formalises this reasoning by stating that probabilities should be assigned by maximizing uncertainty (information entropy) using evidence as constraint. 
This defines a variational problem, where the information entropy functional $H\left[p_1, \dots, p_N\right]$, defined in eq.~\eqref{eq:entropy}, has to be maximized.

The maximisation of entropy, supplemented by evidence in the form of constraints to which the sought-for probability distribution must obey, gives rise to several of the most common probability distributions commonly employed in statistics. For instance, whenever the only constraint available is the normalization of the probability distribution (i.e. no evidence is available), the entropy is maximised by the uniform distribution. If we have evidence to constraint the expected value, the information entropy is maximised by the exponential distribution.

Of particular relevance for our purposes is the case in which, in addition to the mean, also the variance is known: MAXENT leads to the Gaussian distribution. 
This derivation is particularly interesting from the foundational point of view, since it provides a deeper insight into the ubiquitous Gaussian distribution.
Indeed, it is not only the limit distribution provided by the central limit theorem for finite variance processes but it is also the distribution that maximizes the entropy for a fixed mean and variance: from the MAXENT principle, it is the correct probability distribution to assign if mean and covariance are the only quantities that fully define our process. In some sense, we can interpret the central limit theorem as the natural `statistical' evolution toward a configuration that maximizes entropy in repeated experiments.

For this work, we are especially interested in the multi-dimensional case. Suppose we have a vector of measurements $(x(t_1),\ldots,x(t_n)) = (x_1, \ldots, x_n)$ that we conveniently express as a single realization of an unknown stochastic process $x(t)$ and we have information about the expectation value of the process $\mu(t)$ and on the matrix of autocovariances $C_{ij} \equiv C(t_i,t_j)$, then the MAXENT distribution is the $n$-dimensional multivariate Gaussian distribution~\citep{gregory_2005}: 
\begin{align}
    p&\left((x_1, \ldots, x_n)\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det C\right)^{k / 2}}\exp\left(-\frac{1}{2}\sum_{i,j}(x_i-\mu_i) (x_j-\mu_j)C^{-1}_{ij} \right)\,. 
\end{align}

For a wide-sense stationary process the mean function is independent of time, hence it can be redefined to be equal to zero without loss of generality, and the auto-covariance function is dependent only on the time lag $\tau \equiv t_i - t_j$. One can thus choose a sampling rate $\Delta t$ so that $C_{ij} = C((i-j)\Delta t)$. The autocovariance matrix thus becomes a Toeplitz matrix\footnote{
We remind the reader that a Toeplitz matrix is a matrix in the form:
\begin{pmatrix}
	a_0 & a_1 & a_2 & \ldots & \ldots& \ldots &a_n\\
	a_{-1} & a_0 & a_1 & \ldots & \ldots& \ldots &a_{n-1}\\
	a_{-2} & a_{-1} & a_0 & \ldots & \ldots & \ldots  &a_{n-2}\\
	\vdots & \vdots & \vdots & \vdots & \vdots & \vdots &\vdots\\
	a_{-n +1} & \ldots & \ldots & \ldots& a_{-1} & a_0    &a_{1}\\
	a_{-n} & \ldots & \ldots & \ldots& a_{-2} & a_{-1} & a_0
\end{pmatrix}
}.
Toeplitz matrices are asymptotically equivalent to circulant matrices and thus diagonalized by the discrete Fourier transform base~\citep{Gray}.
%R. M. Gray, Foundations and Trends⃝R in Communications and Information Theory 2, 155 (2005), URL https://doi.org/10.1561/0100000006.
Some simple algebra shows that the time-domain multivariate Gaussian can be transformed into the equivalent frequency domain 
probability distribution:
\begin{align}\label{eq:Whittle}
p&\left((\tilde{x}_1, \ldots, \tilde{x}_{n/2})\vert I\right) = \nonumber \\
    &\frac{1}{\left(2 \pi \det S\right)^{n / 2}}\exp\left(-\frac{1}{2}\sum_{ij}\tilde{x}_i S^{-1}_{ij} \tilde{x}_j \right)\,,
\end{align}
where the matrix $S_{ij} = S_i \delta_{ij}$ is an $n\times n$ diagonal matrix whose elements are the PSD $S(f)$ calculated at frequency $f_i$.
Many readers will recognise the familiar form of the Whittle likelihood that stands at the basis of the \emph{matched filter} method\citep{prob_information_theory}
and of gravitational waves data analysis, \citep[e.g.]{Finn_1992, FINDCHIRP}.
Thanks to MAXENT, the problem of defining the probability distribution describing a wide-sense stationary process is thus 
entirely reduced to the estimation of the PSD or, equivalently, the autocovariance function.

@@

In principle, if the autocorrelation was known exactly (i.e. at every time $\tau \in (-\infty,+\infty)$), the computation of the PSD would reduce to a single Fourier transform. However, in any realistic setting, we are dealing with a finite number of samples $N$ from the process, hence such computation is impossible.
Moreover, the error $\sigma_k$ in the estimate of the autocorrelation after $k$ steps increases as $\sigma \sim 1/\sqrt{N - k}$\footnote{
This is easily understood: when computing the autocorrelation at order $k$, only $N-k$ examples of the product $x_t x_{t+k}$ are available and the variance of the average value goes as the inverse of the square root of the points considered.
}, so that only few values for the autocorrelation function can actually be computed reliably.
This bring us the core of the problem: how to give an estimate from partial (and noisy) knowledge of the autocorrelation function? MAXENT can guide us in this task without any a priori assumptions on the unavailable data\footnote{Indeed this is the largest difference with the most common Welch method. The latter assumes that the unknown values of the autocorrelation are $0$. Clearly, this assumption is unjustified and MAXENT is a good way to relax this assumption.}.

@@

Before doing so, there is a technicality to solve: the definition of entropy depends on a probability distribution, not on the PSD.
It can be shown ~\citep[e.g.]{AblesMESA, Bartlett} that the variational problem can be formulated in terms of the power spectral density $S(f)$ alone by considering our signal as the result of the filtering a white noise process using a filter with transfer function $T(f)$ equal to $S(f)$\footnote{
A filter with transfer function $T(f)$ takes in input a time series $x_t$ and outputs a times series $y_t$ such that:
$$T(f) = \frac{\tilde{y}(f)}{\tilde{x}(f)}$$
where $\tilde{x}(f)$ denotes the Fourier transform of $x_t$ (and similarly for $y_t$)
}.
The difference in entropy between the input and the output time series (i.e. the entropy gain) obtained by such filter applied on white noise is:
\begin{equation}\label{eq:EntropyGain}
    \Delta H = \int_{- Ny}^{Ny}\log S(f) df\,.
\end{equation}
where $\Delta t$ is sampling rate and $Ny \equiv \frac{1}{2 \Delta t}$  is the Nyquist frequency.
Thus maximising Eq.~\eqref{eq:EntropyGain} is equivalent to maximizing Eq.~\eqref{eq:entropy_continuos}.
